---
title: "DS3100_Final_Project"
author: "Michael Tung"
date: "2025-04-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Final Project

## 1: Research Question

How do the factors of age, BMI, and daily steps affect quality of sleep? (Linear)

Which factor – physical activity level, stress level, and sleep duration – has the strongest association with developing a sleep disorder? (Logistic)	

## 2: Dataset and Data Wrangling

# Load Packages
```{r}
library(tidyverse)
library(janitor)
library(ggplot2)
library(factoextra)
```

# Load Data and Data Wrangling
```{r}
sleep <- read.csv("/Users/michaeltung/Desktop/DS_3100/DS3100_Final_Project/Sleep_health_and_lifestyle_dataset.csv")

# Load and clean for linear regression section

# Load and clean for logistic regression section

# Load and clean for Advanced Topics Section
sleepPCA <- sleep %>%
  clean_names(case = "small_camel") %>%
  select(sleepDisorder, physicalActivityLevel, stressLevel, sleepDuration) %>%
  drop_na() %>%
  mutate(
    sleepDisorder = as.factor(ifelse(sleepDisorder == "None", 0, 1))
  )
```

## 3: Visualize Variables
```{r}

```

## 4: Regression Models
```{r}

```

## 5: Regression Model Visualization
```{r}

```

## 6: Advanced Topics
```{r}
# PCA
pca_input <- sleepPCA %>% select(-sleepDisorder)

pca_result <- prcomp(pca_input, center = TRUE, scale. = TRUE)

summary(pca_result)

# Clustering
# Calculate % variance explained for PC1 and PC2
pc_var <- summary(pca_result)$importance[2, 1:2]
pc1_var <- round(pc_var[1] * 100, 1)
pc2_var <- round(pc_var[2] * 100, 1)

pca_scores <- as.data.frame(pca_result$x[, 1:2]) # First two principal components

kmeans_result <- kmeans(pca_scores, centers = 3, nstart = 25)

# Add cluster labels to PCA scores
pca_scores$cluster <- as.factor(kmeans_result$cluster)

# Make sure labels are numeric
true_labels <- as.numeric(sleepPCA$sleepDisorder)
predicted_clusters <- as.numeric(kmeans_result$cluster)

# ARI: Adjusted Rand Index
ari_score <- adjustedRandIndex(true_labels, predicted_clusters)

# AMI: Adjusted Mutual Information
# Use aricode package
library(aricode)
ami_score <- AMI(true_labels, predicted_clusters)

# Print results
print(paste("Adjusted Rand Index (ARI):", round(ari_score, 3)))
print(paste("Adjusted Mutual Information (AMI):", round(ami_score, 3)))


```

## 7: Advanced Topic Visualization
```{r}
factoextra::fviz_eig(pca_result)

# Improved PCA Cluster Plot
ggplot(pca_scores, aes(x = PC1, y = PC2, color = cluster)) +
  geom_point(size = 2.5, alpha = 0.8) +
  labs(
    title = "K-Means Clustering of Sleep Patterns",
    x = paste0("Principal Component 1 (", pc1_var, "% variance)"),
    y = paste0("Principal Component 2 (", pc2_var, "% variance)"),
    color = "Cluster"
  ) +
  theme_minimal(base_size = 14) +
  scale_color_brewer(palette = "Set2")

# Plot WSS (Elbow method)
fviz_nbclust(pca_scores[, 1:2], kmeans, method = "wss") +
  labs(title = "Elbow Method for Optimal k (PCA-Reduced Features)")
```

## 8: Conclusion
```{r}

```


