---
title: "DS3100_Final_Project"
author: "Michael Tung, Devanshu Maniar, Paterne Migwaguro"
date: "2025-04-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Final Project

## 1: Research Questions

In this project, we will be investigating the following research questions:

1. Which factor – participants' age, BMI category, or the number of daily steps taken – has the strongest relationship with participants' quality of sleep? (Linear)

2. Which factor – physical activity level, stress level, or sleep duration – has the strongest association with developing a sleep disorder? (Logistic)

## 2: Dataset and Data Wrangling

# Load Packages
```{r}
library(tidyverse)
library(janitor)
library(ggplot2)
library(factoextra)
library(mclust)
library(aricode)
```

# Load Data, Data Overview, and Basic Data Wrangling
```{r}
# Loading the "Sleep_health_and_lifestyle_dataset.csv" file and performing data overview.
sleep <- read.csv("/Users/michaeltung/Desktop/DS_3100/DS3100_Final_Project/Sleep_health_and_lifestyle_dataset.csv")
View(sleep)

lapply(sleep, class)
str(sleep)

dim(sleep)
sapply(sleep, function(x) sum(is.na(x)))


# Load and clean for Linear Regression Section
sleepLinear <- sleep %>%
  clean_names(case = "small_camel") %>%
  select(age, bmiCategory, dailySteps, qualityOfSleep)


# Load and clean for Logistic Regression Section
sleepLogistic <- sleep %>%
  clean_names(case = "small_camel") %>%
  mutate(sleepDisorder = as.factor(ifelse(sleepDisorder == "None", "No", "Yes"))) %>%
  select(physicalActivityLevel, stressLevel, sleepDuration, sleepDisorder)


# Load and clean for Advanced Topics Section
sleepPCA <- sleep %>%
  clean_names(case = "small_camel") %>%
  select(sleepDisorder, physicalActivityLevel, stressLevel, sleepDuration) %>%
  drop_na() %>%
  mutate(
    sleepDisorder = as.factor(ifelse(sleepDisorder == "None", 0, 1))
  )
```

## Discussion of Data

This data was obtained from Kaggle and it presents several helpful variables, all of which are related to sleep. It is unclear where the data first originated from, however, it likely came from a sleep study. Our purpose for the data is to use it to identify our research questions.

Our variables of interest are:

age - the age of a person (continuous)

bmiCategory - BMI of a person (categorical)

dailySteps - Number of steps taken daily by a person (continuous)

qualityOfSleep - Quality of sleep for a person (cateogorical, higher the better)

physicalActivityLevel - Quality of phyical activity a person gets per day (categorical, higher the better)

stressLevel - Stress level of a person (categorical, higher stress level means higher stress)

sleepDuration - Hours of sleep a person gets (continuous)

sleepDisorder - Whether or not a person has a sleep disorder, 1 for yes, 0 for no (Binary)

## 3A: Visualize Variables (Linear Regression)
```{r}
# Visualizing the predictor variable `age` and the response variable `qualityOfSleep`

# `age`: Numeric, continuous variable
# `qualityOfSleep`: Numeric, continuous variable

sleepLinear %>%
ggplot(aes(x = age, y = qualityOfSleep)) +
  geom_jitter(width = 0.2, height = 0.2) +
  geom_smooth(method = "lm") +
  labs(title = "Relating Participants' Age and Sleep Quality",
       x = "Age of Participants",
       y = "Hours of Sleep Received by Participants") +
  theme_minimal()


# Visualizing the predictor variable `bmiCategory` and the response variable
# `qualityOfSleep`

# `bmiCategory`: Character, categorical variable
# `qualityOfSleep`: Numeric, continuous variable

sleepLinear %>%
ggplot(aes(x = bmiCategory, y = qualityOfSleep)) +
  geom_boxplot() +
  labs(title = "Relating Participants' BMI Category and Sleep Quality",
       x = "BMI Category of Participants",
       y = "Hours of Sleep Received by Participants") +
  theme_minimal()


# Visualizing the predictor variable `dailySteps` and the response variable
# `qualityOfSleep`

# `Daily Steps`: Numeric, continuous variable
# `Quality of Sleep`: Numeric, continuous variable

sleepLinear %>%
ggplot(aes(x = dailySteps, y = qualityOfSleep)) +
  geom_jitter(width = 0.2, height = 0.2) +
  geom_smooth(method = "lm") +
  labs(title = "Relating Participants' Daily Steps and Sleep Quality",
       x = "Daily Average Number of Steps Taken by Participants",
       y = "Hours of Sleep Received by Participants") +
  theme_minimal()
```

## 3B: Visualize Variables (Logistic Regression)
```{r}
# Visualizing the predictor variable `physicalActivityLevel` and the response variable
# `sleepDisorder`

# `physicalActivityLevel`: Numeric, categorical variable (since these represent levels)
# `sleepDisorder`: Factor, categorical variable

sleepLogistic %>%
ggplot(aes(x = physicalActivityLevel, y = sleepDisorder)) +
  geom_bar(stat = "identity") +
  labs(title = "Relating Participants' Physical Activity Levels with Sleep Disorder",
       x = "Activity Level of Participants",
       y = "Whether or Not Participants Have a Sleep Disorder") +
  theme_minimal()


# Visualizing the predictor variable `stressLevel` and the response variable
# `sleepDisorder`

# `stressLevel`: Numeric, categorical variable (since these represent levels)
# `sleepDisorder`: Factor, categorical variable

sleepLogistic %>%
ggplot(aes(x = stressLevel, y = sleepDisorder)) +
  geom_bar(stat = "identity") +
  labs(title = "Relating Participants' Stress Levels with Sleep Disorder",
       x = "Stress Level of Participants",
       y = "Whether or Not Participants Have a Sleep Disorder") +
  theme_minimal()


# Visualizing the predictor variable `sleepDuration` and the response variable
# `sleepDisorder`

# `sleepDuration`: Numeric, continuous variable
# `sleepDisorder`: Factor, categorical variable

sleepLogistic %>%
ggplot(aes(x = sleepDuration, y = sleepDisorder)) +
  geom_boxplot() +
  labs(title = "Relating Participants' Sleep Duration with Sleep Disorder",
       x = "Sleep Duration (in Hours) of Participants",
       y = "Whether or Not Participants Have a Sleep Disorder") +
  theme_minimal()
```

## 4A: Regression Model (Linear Regression)
```{r}
# Running a multivariate linear regression on participants' quality of sleep given their
# age, BMI category, and daily average number of steps taken.
lm_age_bmi_daily_steps_sleep <- lm(formula = qualityOfSleep ~ age + bmiCategory +
                                     dailySteps,
                                   data = sleepLinear)


# Printing the output of the above multivariate linear regression model.
summary(lm_age_bmi_daily_steps_sleep)
modelsummary(lm_age_bmi_daily_steps_sleep, output = "markdown")


# Performing the Shapiro - Wilk test on the residuals (or error terms) of the
# multivariate linear regression model to investigate normality.
shapiro.test(residuals(lm_age_bmi_daily_steps_sleep))


# Plotting each of the independent variables separately with the multivariate linear
# regression model's residuals to closely investigate which variable(s) is / are
# contributing to potential non - linearity, non - normality, and / or
# heteroscedasticity. These make up our residual plots.
plot(sleepLinear$age, residuals(lm_age_bmi_daily_steps_sleep))

sleepLinear$bmiCategory <- factor(sleepLinear$bmiCategory)
boxplot(residuals(lm_age_bmi_daily_steps_sleep) ~ sleepLinear$bmiCategory)

plot(sleepLinear$dailySteps, residuals(lm_age_bmi_daily_steps_sleep))
```

## 4B: Regression Model (Logistic Regression)
```{r}
# Performing a logistic binomial regression to predict the log - likelihood of the
# participants developing a sleep disorder given their physical activity level, stress
# level, and sleep duration (in hours).
logistic_binom_sleep_disorder <- glm(formula = sleepDisorder ~ physicalActivityLevel + stressLevel 
                                     + sleepDuration,
                                     data = sleepLogistic,
                                     family = "binomial")


# Printing the output of the above logistic binomial regression model.
summary(logistic_binom_sleep_disorder, output = "markdown")
```

## 5A: Regression Model Visualization (Linear Regression)
```{r}
# Plotting a residuals vs. fitted values plot between the multivariate linear model and
# its residuals (or error terms) to investigate linearity and homoscedasticity.
plot(lm_age_bmi_daily_steps_sleep, which = 1)


# Plotting a QQ plot (or a probability plot) between the multivariate linear model 
# (i.e. theoretical quantiles) and its standardized residuals (or error terms) to
# investigate normality.
plot(lm_age_bmi_daily_steps_sleep, which = 2)


# Plotting a histogram of the distribution of the multivariate linear model's residuals
# (or error terms) to investigate normality.
sleepLinear %>%
  ggplot(aes(x = residuals(lm_age_bmi_daily_steps_sleep))) +
  geom_histogram(bins = 8) +
  labs(title = "The Distribution of Participants' Linear Model Residuals",
       x = "Multivariate Linear Regression Model's Residuals",
       y = "Count of the Multivariate Regression Model's Residuals")
```

## 5B: Regression Model Visualization (Logistic Regression)
```{r}
# Creating an array of the predictor variables (and intercept) used in the logistic
# binomial model on the participants. This prints out each variables' `Estimate`,
# `Std. Error`, `z value`, and `Pr(>|z|)` in a tabular format.
coef_summary <- summary(logistic_binom_sleep_disorder)$coefficients


# Transforming the predictor variables array to a data frame, and making some of the
# column names more specific.
coef_df <- data.frame(
  Variable = rownames(coef_summary),
  Estimate = coef_summary[, "Estimate"],
  Standard_Error = coef_summary[, "Std. Error"]
)


# Using the predictor variables data frame to create a bar plot illustrating confidence
# intervals on each of the light green variables' estimates (including the intercept).
coef_df %>%
  ggplot(aes(x = Variable, y = Estimate)) +
  geom_col(fill = "lightgreen") +
  geom_errorbar(aes(ymin = Estimate - Standard_Error, ymax = Estimate + Standard_Error),
                width = 0.2) +
  coord_flip() +
  labs(title = "Coefficient Estimates on the Participants", 
       subtitle = "Physical Activity, Stress, and Sleep Duration on Sleep Disorder",
       x = "Variable",
       y = "Coefficient Estimate") +
  theme_minimal()
```

## 6: Correlation Analyses for **Linear Regression**
```{r}
# Performing a Pearson correlation between the numeric variables `age` and participants'
# quality of sleep.
age_quality_sleep_pearson <- cor(sleepLinear$age,
                                 sleepLinear$qualityOfSleep,
                                 method = "pearson")
print(age_quality_sleep_pearson)


# Performing Spearman's rank correlation between the categorical variable `bmiCategory`
# and participants' quality of sleep (numeric variable).
sleepLinear$bmiCategory <- as.numeric(sleepLinear$bmiCategory)

bmi_quality_sleep_spearman <- cor(sleepLinear$bmiCategory,
                                  sleepLinear$qualityOfSleep,
                                  method = "spearman")
print(bmi_quality_sleep_spearman)


# Performing a Pearson correlation between the numeric variables `dailySteps` and
# participants' quality of sleep.
daily_steps_quality_sleep_pearson <- cor(sleepLinear$dailySteps,
                                         sleepLinear$qualityOfSleep,
                                         method = "pearson")
print(daily_steps_quality_sleep_pearson)
```

## 7: Advanced Topics
```{r}
# PCA
pca_input <- sleepPCA %>% select(-sleepDisorder)

pca_result <- prcomp(pca_input, center = TRUE, scale. = TRUE)

summary(pca_result)

# Clustering
# Calculate % variance explained for PC1 and PC2
pc_var <- summary(pca_result)$importance[2, 1:2]
pc1_var <- round(pc_var[1] * 100, 1)
pc2_var <- round(pc_var[2] * 100, 1)

pca_scores <- as.data.frame(pca_result$x[, 1:2]) # First two principal components

print("Estimating PCA Categories")
pca_result$rotation[, 1:2]

kmeans_result <- kmeans(pca_scores, centers = 2, nstart = 25)

# Add cluster labels to PCA scores
pca_scores$cluster <- as.factor(kmeans_result$cluster)

# Make sure labels are numeric
true_labels <- as.numeric(sleepPCA$sleepDisorder)
predicted_clusters <- as.numeric(kmeans_result$cluster)

# ARI: Adjusted Rand Index
ari_score <- adjustedRandIndex(true_labels, predicted_clusters)

# AMI: Adjusted Mutual Information
ami_score <- AMI(true_labels, predicted_clusters)

# Print results
print(paste("Adjusted Rand Index (ARI):", round(ari_score, 3)))
print(paste("Adjusted Mutual Information (AMI):", round(ami_score, 3)))
```

## 8: Advanced Topic Visualization
```{r}
factoextra::fviz_eig(pca_result)

# Plot WSS (Elbow method) to see ideal K for k-means clustering
fviz_nbclust(pca_scores[, 1:2], kmeans, method = "wss") +
  labs(title = "Elbow Method for Optimal k (PCA-Reduced Features)")

# Improved PCA Cluster Plot
ggplot(pca_scores, aes(x = PC1, y = PC2, color = cluster)) +
  geom_point(size = 2.5, alpha = 0.8) +
  labs(
    title = "K-Means Clustering of Sleep Patterns",
    x = paste0("Principal Component 1 (", pc1_var, "% variance)"),
    y = paste0("Principal Component 2 (", pc2_var, "% variance)"),
    color = "Cluster"
  ) +
  theme_minimal(base_size = 14) +
  scale_color_brewer(palette = "Set2")
```

## 9: Conclusions

Linear and Logistic Regression Conclusion:

With regards to the research question for linear regression, the data visualizations make it clear that participants' `age` has the greatest linear correlation with the hours of sleep participants received (i.e. `qualityOfSleep`) rather than `bmiCategory` and `dailySteps`. This is because the linear model has the greatest, positive slope for `age` in that as the age of the participants increases, their sleep quality increases. Intuitively, this makes sense because older participants likely become more attentive to their self-care needs and prioritize getting the recommended amount of sleep (8-9 hours) so that a lack of sleep doesn't disrupt their productivity at work. Although the graph doesn't illustrate this, it's not surprising that those in their 20s get roughly no more than 6 hours of sleep. People in their 20s are completing either their associate's or bachelor's degree, or even pursuing their master's or PhDs. Not only do these heavy university courseloads can definitely take up most of their time, college students commonly have stress and anxiety about grades/future careers which can lead to difficulty sleeping. Additionally, taking earning morning college courses and being involved in late-night study sessions and/or club organizations can clearly disrupt sleep consistency. It appears that the variable that has no evident correlation with `qualityOfSleep` is `dailySteps` since its linear model is a horizontal line indicating no trend/pattern/relationship.

With regards to the research question for logistic regression, the data visualizations give us some belief that `sleepDuration` has the greatest impact on `sleepDisorder`. The boxplots that illustrate this relationship show us that the median value (indicated by the dark black, vertical line) is significantly lower for participants who have a sleep disorder (which is at about `6.5` hours) than for participants who don't have a sleep disorder (which is at about `7.5` hours). Again, this intuitively makes sense because sleep disorders like `Sleep Apnea` and `Insomnia` make it difficult for people to fall asleep at night, and this consequently disrupts their sleep duration (or the amount of sleep these participants get). The variable `stressLevel` appears to have an unexpected inverse relationship on sleep (i.e. the greater the stress level, it's more likely that these participants don't have a sleep disorder and vice versa). For `physicalActivityLevel`, the relationship appears to be something that we would expect; participants who engage in more physical activity are likely to not have a sleep disorder and vice versa. The problem is that since we used different types of plots and "eyeballed" our conclusions, we would need a more precise metric like linear and logistic regression models to affirm our claims.

When I ran a multivariable linear regression model on the predictor variables `age`, `bmiCategory`, and `dailySteps` on the response variable `qualityOfSleep`, I noticed that the `Pr(>|t|)` value for `dailySteps` is `0.237`. Since this value is a lot greater than the alpha value of `0.05`, we don't have enough evidence to conclude that `dailySteps` has a clear, linear relationship with `qualityOfSleep`. This affirms the insights we made earlier as the linear model for this relationship showed a horizontal line implying no actual correlation. As expected, we see that the `Pr(>|t|)` values for the other variables `age` and `bmiCategory` are lower than `0.05` which suggests that it's not likely that these variables are correlated with `qualityOfSleep` **as a result of chance/luck**. Additionally, the `R2` and `R2 Adj.` values are `0.643` and `0.640` which is moderate - it's the best nor the worst. This isn't surprising because we have a variable (`dailySteps`) that doesn't have a significant relationship with participants' quality of sleep, so this could be the reason why the overall `R2` and `R2 Adj.` values are low. As a note for the future, we can certainly remove the `dailySteps` variable and likely see greater variation values in `age` and `bmiCategory` as explained by the model (`qualityOfSleep`). But in this case, these average values imply a moderate linear relationship, provided that higher R-squared values are indicative of a greater model fit.

For reference, the estimated linear model is shown below:

$$
QualityofSleep = 3.934 + \sum_{i=1}^{n} (age_i * 0.1188 + bmiCategory_i * -0.6297 + dailySteps_i * -0.00002732) + \epsilon
$$
Since this is a linear regression model, I decided to run a Shapiro-Wilk test to check if the model is normally distributed and not skewed in any particular direction. The `W` or normality value that we get from the model is `0.9699` which suggests that our model appears to be normal, but by checking that the `p-value` (`5.496e-07`) is less than `0.05`, we must reject the null hypothesis and conclude that the model isn't normally distributed. As another note for the future, we can apply transformation to either one of the independent and/or dependent variable(s) to normalize the residuals. To investigate this issue further, I decided to plot the response variable with each of the independent variables separately using a residual plot to check for potential non-linearity (i.e. no clear trend/pattern) and potential heteroscedasticity (i.e. no funnel shape created by the distribution of the residuals). While the residual plots appear to show linearity and homoscedasticity for the variables `age` and `dailySteps`, we see that `bmiCategory` doesn't satisfy linearity. Specifically, we see that the median values for the boxplots create a "arch" shape among the `Normal`, `Normal Weight`, `Obese`, and `Overweight` categories. This is obviously a clear pattern we should note of, and appropriately apply the necessary transformations on this variable in specific to further improve this linear regression model. The fact that this model is homoscedastic because the residuals vs. fitted lines plot for the overall model appears to have no clear trend/pattern or funnel shape. As expected from the Shapiro-Wilk test, the probability plot that illustrates normality shows that the model isn't normally distributed; the residuals aren't largely situated on the line of normality (which is the light-gray, dotted line). Not to mention, the correlation analyses easily show that `age` has the largest absolute relationship with `qualityOfSleep` at `0.4737339`.

When I ran a logistic binomial regression model on the predictor variables `physicalActivityLevel`, `stressLevel`, and `sleepDuration` on the response variable `sleepDisorder`, all of the predictor variables' `Pr(>|t)` values are less than `0.05`. This suggests that all of these variables have some type of correlation with `sleepDisorder` that's **not due to chance/luck**. Overall, I would say that this model generally seeks to find fit with the response variable. I say this because the residual deviance (i.e. which explains model fit) is `432.90` and the degrees of freedom is `370`. Since `432.90` / `370` is about `1.17`, we have some overdispersion (i.e. extreme variability/outliers) in this case. As another note for the future, we could use the quasi-Poisson model to account for this extra variability.

For reference, the estimated logistic binomial model is shown below:

$$
P(default = 1 | physicalActivityLevel, stressLevel, sleepDuration) = \frac{1}{1 + e^{-(15.6398 + \sum_{i=1}^{n} physicalActivityLevel_i * 0.02909 + stressLevel_i * -0.4761 + sleepDuration_i * -2.1380 + \epsilon)}}
$$
The `Estimate` values for `physicalActivityLevel`, `stressLevel`, and `sleepDuration` are the following, respectively: `0.029088`, `-0.476137`, and `-2.137999`. So, the odds ratio would be the following, respectively: `1.029515`, `0.6211784`, and `0.1178905`. These odds ratio contribute to roughly a `2.9515`% **increase** in the likelihood of having a sleep disorder, a `37.88216`% **decrease** in the likelihood of having a sleep disorder, and an `88.21095`% **decrease** in the likelihood of having a sleep disorder, respectively. This is clearly evident in the coefficient estimates visualization for each of the independent variables on `sleepDisorder`. The larger the green bar (when comparing the independent variables at least), the more likely that independent variable has a correlation with `sleepDisorder`. This plot shows us that `sleepDuration` is indeed this variable.

Overall, to answer the research question, we have the following:

(Linear) Among participants' age, BMI category, and the number of daily steps taken, the variable that has the strongest relationship with participants' quality of sleep is `age`. This conclusion is made after taking into account model homoscedasticity, a clear, positive trend from its linear model, and model normality that this variable satisfies.

(Logistic) Among participants' physical activity level, stress level, or sleep duration, the variable that has the strongest association with developing a sleep disorder is `sleepDuration` as shown by the greatest change in the odds ratio and its large coefficient estimate relative to any other predictor variable.

Clustering and PCA conclusion:

The PCA results demonstrated that roughly two components can explain 94% of the variation in the data (between the three variables physicalActivityLevel, stressLevel, sleepDuration). The result of the PCA was plotted using the Scree plot where we see dimension 3 is significantly lower than dimensions 1 and 2. Using the rotation matrix of the pca results, we see that PC1 is likely quality of sleep, since it is most explained by sleep duration, and PC2 is likely physical activity level, since it is most explained by physicalActivityLevel. Since the Adjusted Rand Index and Adjusted Mutual Information are relatively low, the clusters derived from behavioral data do not strongly align with diagnosed sleep disorders. From the PCA's results, it can be shown that quality of sleep likely has the highest relationship with developing sleep disorders, however, there is not a strong allignment/fit.