---
title: "DS3100_Final_Project"
author: "Michael Tung"
date: "2025-04-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Final Project

## 1: Research Question

How do the factors of age, BMI, and daily steps affect quality of sleep? (Linear)

Which factor – physical activity level, stress level, and sleep duration – has the strongest association with developing a sleep disorder? (Logistic)	

## 2: Dataset and Data Wrangling

# Load Packages
```{r}
library(tidyverse)
library(janitor)
library(ggplot2)
library(factoextra)
```

# Load Data and Data Wrangling
```{r}
sleep <- read.csv("/Users/michaeltung/Desktop/DS_3100/DS3100_Final_Project/Sleep_health_and_lifestyle_dataset.csv")

# Load and clean for linear regression section

# Load and clean for logistic regression section

# Load and clean for Advanced Topics Section
sleepPCA <- sleep %>%
  clean_names(case = "small_camel") %>%
  select(sleepDisorder, physicalActivityLevel, stressLevel, sleepDuration) %>%
  drop_na() %>%
  mutate(
    sleepDisorder = as.factor(ifelse(sleepDisorder == "None", 0, 1))
  )
```

## 3: Visualize Variables
```{r}

```

## 4: Regression Models
```{r}

```

## 5: Regression Model Visualization
```{r}

```

## 6: Advanced Topics
```{r}
# PCA
pca_input <- sleepPCA %>% select(-sleepDisorder)

pca_result <- prcomp(pca_input, center = TRUE, scale. = TRUE)

summary(pca_result)

# Clustering
```

## 7: Advanced Topic Visualization
```{r}
factoextra::fviz_eig(pca_result)
```

## 8: Conclusion
```{r}

```


