---
title: "DS3100_Final_Project"
author: "Michael Tung, Devanshu Maniar, Paterne Migwaguro"
date: "2025-04-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Final Project

## 1: Research Question

How do the factors of age, BMI, and daily steps affect quality of sleep? (Linear)

Which factor – physical activity level, stress level, and sleep duration – has the strongest association with developing a sleep disorder? (Logistic)	

## 2: Dataset and Data Wrangling

# Load Packages
```{r}
library(tidyverse)
library(janitor)
library(ggplot2)
library(factoextra)
library(mclust)
library(aricode)
```

# Load Data and Data Wrangling
```{r}
sleep <- read.csv("/Users/michaeltung/Desktop/DS_3100/DS3100_Final_Project/Sleep_health_and_lifestyle_dataset.csv")

# Load and clean for linear regression section

# Load and clean for logistic regression section

# Load and clean for Advanced Topics Section
sleepPCA <- sleep %>%
  clean_names(case = "small_camel") %>%
  select(sleepDisorder, physicalActivityLevel, stressLevel, sleepDuration) %>%
  drop_na() %>%
  mutate(
    sleepDisorder = as.factor(ifelse(sleepDisorder == "None", 0, 1))
  )
```

## 3: Visualize Variables
```{r}

```

## 4: Regression Models
```{r}

```

## 5: Regression Model Visualization
```{r}

```

## 6: Advanced Topics
```{r}
# PCA
pca_input <- sleepPCA %>% select(-sleepDisorder)

pca_result <- prcomp(pca_input, center = TRUE, scale. = TRUE)

summary(pca_result)

# Clustering
# Calculate % variance explained for PC1 and PC2
pc_var <- summary(pca_result)$importance[2, 1:2]
pc1_var <- round(pc_var[1] * 100, 1)
pc2_var <- round(pc_var[2] * 100, 1)

pca_scores <- as.data.frame(pca_result$x[, 1:2]) # First two principal components

print("Estimating PCA Categories")
pca_result$rotation[, 1:2]

kmeans_result <- kmeans(pca_scores, centers = 2, nstart = 25)

# Add cluster labels to PCA scores
pca_scores$cluster <- as.factor(kmeans_result$cluster)

# Make sure labels are numeric
true_labels <- as.numeric(sleepPCA$sleepDisorder)
predicted_clusters <- as.numeric(kmeans_result$cluster)

# ARI: Adjusted Rand Index
ari_score <- adjustedRandIndex(true_labels, predicted_clusters)

# AMI: Adjusted Mutual Information
ami_score <- AMI(true_labels, predicted_clusters)

# Print results
print(paste("Adjusted Rand Index (ARI):", round(ari_score, 3)))
print(paste("Adjusted Mutual Information (AMI):", round(ami_score, 3)))
```

## 7: Advanced Topic Visualization
```{r}
factoextra::fviz_eig(pca_result)

# Plot WSS (Elbow method) to see ideal K for k-means clustering
fviz_nbclust(pca_scores[, 1:2], kmeans, method = "wss") +
  labs(title = "Elbow Method for Optimal k (PCA-Reduced Features)")

# Improved PCA Cluster Plot
ggplot(pca_scores, aes(x = PC1, y = PC2, color = cluster)) +
  geom_point(size = 2.5, alpha = 0.8) +
  labs(
    title = "K-Means Clustering of Sleep Patterns",
    x = paste0("Principal Component 1 (", pc1_var, "% variance)"),
    y = paste0("Principal Component 2 (", pc2_var, "% variance)"),
    color = "Cluster"
  ) +
  theme_minimal(base_size = 14) +
  scale_color_brewer(palette = "Set2")
```

## 8: Conclusion

Clustering and PCA conclusion:

The PCA results demonstrated that roughly two components can explain 94% of the variation in the data (between the three variables physicalActivityLevel, stressLevel, sleepDuration). The result of the PCA was plotted using the Scree plot where we see dimension 3 is significantly lower than dimensions 1 and 2. Using the rotation matrix of the pca results, we see that PC1 is likely quality of sleep, since it is most explained by sleep duration, and PC2 is likely physical activity level, since it is most explained by physicalActivityLevel. Since the Adjusted Rand Index and Adjusted Mutual Information are relatively low, the clusters derived from behavioral data do not strongly align with diagnosed sleep disorders.


