---
title: "DS3100_Final_Project"
author: "Michael Tung, Devanshu Maniar, Paterne Migwaguro"
date: "2025-04-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Final Project

## 1: Research Questions

In this project, we will be investigating the following research questions:

1. Which factor – participants' age, BMI category, or the number of daily steps taken – has the strongest relationship with participants' quality of sleep? (Linear)

2. Which factor – physical activity level, stress level, or sleep duration – has the strongest association with developing a sleep disorder? (Logistic)

## 2: Dataset and Data Wrangling

# Load Packages
```{r}
library(janitor)
library(ggplot2)
library(factoextra)
library(mclust)
library(aricode)
library(MASS)
library(tidyverse)
library(ggeffects)
library(nnet)
library(patchwork)
```

# Load Data, Data Overview, and Basic Data Wrangling
```{r}
# Loading the "Sleep_health_and_lifestyle_dataset.csv" file and performing data overview.
sleep <- read.csv("Sleep_health_and_lifestyle_dataset.csv")
#View(sleep)

lapply(sleep, class)
#str(sleep)

dim(sleep)
sapply(sleep, function(x) sum(is.na(x)))


# Load and clean for Linear Regression Section
sleepLinear <- sleep %>%
  clean_names(case = "small_camel") %>%
 select(age, bmiCategory, dailySteps, qualityOfSleep)


# Load and clean for Logistic Regression Section
sleepLogistic <- sleep %>%
  clean_names(case = "small_camel") %>%
  mutate(sleepDisorder = as.factor(ifelse(sleepDisorder == "None", "No", "Yes"))) %>%
  select(physicalActivityLevel, stressLevel, sleepDuration, sleepDisorder)

# Load and clean for Multinomial Logistic Regression Section
sleepMLR <- sleep %>% 
  clean_names(case = "small_camel") %>%
  select(sleepDisorder, physicalActivityLevel, stressLevel, sleepDuration) %>%
  mutate(
    sleepDisorderMulti = case_when(
      str_detect(sleepDisorder, regex("apnea", ignore_case = TRUE)) ~ "Sleep Apnea",
      str_detect(sleepDisorder, regex("insomnia", ignore_case = TRUE)) ~ "Insomnia",
      str_detect(sleepDisorder, regex("none", ignore_case = TRUE)) ~ "None",
      TRUE ~ "Other"
    ),
    sleepDisorderMulti = factor(sleepDisorderMulti, levels = c("None", "Insomnia", "Sleep Apnea", "Other"))
  ) %>% 
  drop_na()

# Load and clean for Ordered Logistic Regression Section
sleepOLR <- sleep %>%
  clean_names(case = "small_camel") %>%
  select(bmiCategory, dailySteps, qualityOfSleep, occupation, gender, age) %>%
  mutate(
    qualityOfSleep = ordered(qualityOfSleep),
    bmiCategory = as.factor(bmiCategory),
    occupation = as.factor(occupation),
    gender = as.factor(gender)
  ) %>%
  mutate(occupation = fct_lump_min(occupation, min = 10, other_level = "Other")) %>%
  mutate(
    age = scale(age),
    dailySteps = scale(dailySteps)
  ) %>%
  drop_na()


# Load and clean for Advanced Topics Section
sleepPCA <- sleep %>%
  clean_names(case = "small_camel") %>%
  select(sleepDisorder, physicalActivityLevel, stressLevel, sleepDuration) %>%
  drop_na() %>%
  mutate(
    sleepDisorder = as.factor(ifelse(sleepDisorder == "None", 0, 1))
  )


```

## 3A: Visualize Variables (Linear Regression)
```{r}
# Visualizing the predictor variable `age` and the response variable `qualityOfSleep`

# `age`: Numeric, continuous variable
# `qualityOfSleep`: Numeric, continuous variable

sleepLinear %>%
ggplot(aes(x = age, y = qualityOfSleep)) +
  geom_jitter(width = 0.2, height = 0.2) +
  geom_smooth(method = "lm") +
  labs(title = "Relating Participants' Age and Sleep Quality",
       x = "Age of Participants",
       y = "Hours of Sleep Received by Participants") +
  theme_minimal()


# Visualizing the predictor variable `bmiCategory` and the response variable
# `qualityOfSleep`

# `bmiCategory`: Character, categorical variable
# `qualityOfSleep`: Numeric, continuous variable

sleepLinear %>%
ggplot(aes(x = bmiCategory, y = qualityOfSleep)) +
  geom_boxplot() +
  labs(title = "Relating Participants' BMI Category and Sleep Quality",
       x = "BMI Category of Participants",
       y = "Hours of Sleep Received by Participants") +
  theme_minimal()


# Visualizing the predictor variable `dailySteps` and the response variable
# `qualityOfSleep`

# `Daily Steps`: Numeric, continuous variable
# `Quality of Sleep`: Numeric, continuous variable

sleepLinear %>%
ggplot(aes(x = dailySteps, y = qualityOfSleep)) +
  geom_jitter(width = 0.2, height = 0.2) +
  geom_smooth(method = "lm") +
  labs(title = "Relating Participants' Daily Steps and Sleep Quality",
       x = "Daily Average Number of Steps Taken by Participants",
       y = "Hours of Sleep Received by Participants") +
  theme_minimal()
```

## 3B: Visualize Variables (Logistic Regression)
```{r}
# Visualizing the predictor variable `physicalActivityLevel` and the response variable
# `sleepDisorder`

# `physicalActivityLevel`: Numeric, categorical variable (since these represent levels)
# `sleepDisorder`: Factor, categorical variable

sleepLogistic %>%
ggplot(aes(x = physicalActivityLevel, y = sleepDisorder)) +
  geom_bar(stat = "identity") +
  labs(title = "Relating Participants' Physical Activity Levels with Sleep Disorder",
       x = "Activity Level of Participants",
       y = "Whether or Not Participants Have a Sleep Disorder") +
  theme_minimal()


# Visualizing the predictor variable `stressLevel` and the response variable
# `sleepDisorder`

# `stressLevel`: Numeric, categorical variable (since these represent levels)
# `sleepDisorder`: Factor, categorical variable

sleepLogistic %>%
ggplot(aes(x = stressLevel, y = sleepDisorder)) +
  geom_bar(stat = "identity") +
  labs(title = "Relating Participants' Stress Levels with Sleep Disorder",
       x = "Stress Level of Participants",
       y = "Whether or Not Participants Have a Sleep Disorder") +
  theme_minimal()


# Visualizing the predictor variable `sleepDuration` and the response variable
# `sleepDisorder`

# `sleepDuration`: Numeric, continuous variable
# `sleepDisorder`: Factor, categorical variable

sleepLogistic %>%
ggplot(aes(x = sleepDuration, y = sleepDisorder)) +
  geom_boxplot() +
  labs(title = "Relating Participants' Sleep Duration with Sleep Disorder",
       x = "Sleep Duration (in Hours) of Participants",
       y = "Whether or Not Participants Have a Sleep Disorder") +
  theme_minimal()
```

## 4A: Regression Model (Linear Regression)
```{r}
# Running a multivariate linear regression on participants' quality of sleep given their
# age, BMI category, and daily average number of steps taken.
lm_age_bmi_daily_steps_sleep <- lm(formula = qualityOfSleep ~ age + bmiCategory +
                                     dailySteps,
                                   data = sleepLinear)


# Printing the output of the above multivariate linear regression model.
summary(lm_age_bmi_daily_steps_sleep)
modelsummary(lm_age_bmi_daily_steps_sleep, output = "markdown")


# Performing the Shapiro - Wilk test on the residuals (or error terms) of the
# multivariate linear regression model to investigate normality.
shapiro.test(residuals(lm_age_bmi_daily_steps_sleep))


# Plotting each of the independent variables separately with the multivariate linear
# regression model's residuals to closely investigate which variable(s) is / are
# contributing to potential non - linearity, non - normality, and / or
# heteroscedasticity. These make up our residual plots.
plot(sleepLinear$age, residuals(lm_age_bmi_daily_steps_sleep))

sleepLinear$bmiCategory <- factor(sleepLinear$bmiCategory)
boxplot(residuals(lm_age_bmi_daily_steps_sleep) ~ sleepLinear$bmiCategory)

plot(sleepLinear$dailySteps, residuals(lm_age_bmi_daily_steps_sleep))
```

## 4B: Regression Model (Logistic Regression)
```{r}
# Performing a logistic binomial regression to predict the log - likelihood of the
# participants developing a sleep disorder given their physical activity level, stress
# level, and sleep duration (in hours).
logistic_binom_sleep_disorder <- glm(formula = sleepDisorder ~ physicalActivityLevel + stressLevel 
                                     + sleepDuration,
                                     data = sleepLogistic,
                                     family = "binomial")


# Printing the output of the above logistic binomial regression model.
summary(logistic_binom_sleep_disorder, output = "markdown")
```

## 5A: Regression Model Visualization (Linear Regression)
```{r}
# Plotting a residuals vs. fitted values plot between the multivariate linear model and
# its residuals (or error terms) to investigate linearity and homoscedasticity.
plot(lm_age_bmi_daily_steps_sleep, which = 1)


# Plotting a QQ plot (or a probability plot) between the multivariate linear model 
# (i.e. theoretical quantiles) and its standardized residuals (or error terms) to
# investigate normality.
plot(lm_age_bmi_daily_steps_sleep, which = 2)


# Plotting a histogram of the distribution of the multivariate linear model's residuals
# (or error terms) to investigate normality.
sleepLinear %>%
  ggplot(aes(x = residuals(lm_age_bmi_daily_steps_sleep))) +
  geom_histogram(bins = 8) +
  labs(title = "The Distribution of Participants' Linear Model Residuals",
       x = "Multivariate Linear Regression Model's Residuals",
       y = "Count of the Multivariate Regression Model's Residuals")
```

## 5B: Regression Model Visualization (Logistic Regression)
```{r}
# Creating an array of the predictor variables (and intercept) used in the logistic
# binomial model on the participants. This prints out each variables' `Estimate`,
# `Std. Error`, `z value`, and `Pr(>|z|)` in a tabular format.
coef_summary <- summary(logistic_binom_sleep_disorder)$coefficients


# Transforming the predictor variables array to a data frame, and making some of the
# column names more specific.
coef_df <- data.frame(
  Variable = rownames(coef_summary),
  Estimate = coef_summary[, "Estimate"],
  Standard_Error = coef_summary[, "Std. Error"]
)


# Using the predictor variables data frame to create a bar plot illustrating confidence
# intervals on each of the light green variables' estimates (including the intercept).
coef_df %>%
  ggplot(aes(x = Variable, y = Estimate)) +
  geom_col(fill = "lightgreen") +
  geom_errorbar(aes(ymin = Estimate - Standard_Error, ymax = Estimate + Standard_Error),
                width = 0.2) +
  coord_flip() +
  labs(title = "Coefficient Estimates on the Participants", 
       subtitle = "Physical Activity, Stress, and Sleep Duration on Sleep Disorder",
       x = "Variable",
       y = "Coefficient Estimate") +
  theme_minimal()
```

## 6: Correlation Analyses for **Linear Regression**
```{r}
# Performing a Pearson correlation between the numeric variables `age` and participants'
# quality of sleep.
age_quality_sleep_pearson <- cor(sleepLinear$age,
                                 sleepLinear$qualityOfSleep,
                                 method = "pearson")
print(age_quality_sleep_pearson)


# Performing Spearman's rank correlation between the categorical variable `bmiCategory`
# and participants' quality of sleep (numeric variable).
sleepLinear$bmiCategory <- as.numeric(sleepLinear$bmiCategory)

bmi_quality_sleep_spearman <- cor(sleepLinear$bmiCategory,
                                  sleepLinear$qualityOfSleep,
                                  method = "spearman")
print(bmi_quality_sleep_spearman)


# Performing a Pearson correlation between the numeric variables `dailySteps` and
# participants' quality of sleep.
daily_steps_quality_sleep_pearson <- cor(sleepLinear$dailySteps,
                                         sleepLinear$qualityOfSleep,
                                         method = "pearson")
print(daily_steps_quality_sleep_pearson)
```


## 7: Multinomial Logistic Regression For Sleep Disorder
```{r}
# Fit the model 
multi_logit <- multinom(
  sleepDisorderMulti ~ physicalActivityLevel + stressLevel + sleepDuration,
  data = sleepMLR
)

multi_logit_summary <- tidy(multi_logit, conf.int = TRUE)
print(multi_logit_summary)
```

### Multinomial Logistic Regression Visualization
```{r}
# Filter out intercepts for plotting
coef_plot_df <- multi_logit_summary %>%
  filter(term != "(Intercept)") %>%
  mutate(
    term = recode(term,
                  physicalActivityLevel = "Physical Activity Level",
                  stressLevel = "Stress Level",
                  sleepDuration = "Sleep Duration")
  )

# Plot: Separate bars for each disorder category vs. None
ggplot(coef_plot_df, aes(x = estimate, y = term, fill = y.level)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2, 
                 position = position_dodge(width = 0.7)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  labs(
    title = "Multinomial Logit Coefficients: Sleep Disorder Types",
    x = "Coefficient Estimate",
    y = "Predictor",
    fill = "Disorder vs. None"
  ) +
  theme_minimal(base_size = 14)


# 1. Predicted probabilities by Stress Level
pred_probs_stress <- ggeffect(multi_logit, terms = "stressLevel")

p1 <- ggplot(pred_probs_stress, aes(x = x, y = predicted, color = group, fill = group)) +
  geom_line(size = 1.2) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.15, color = NA) +
  labs(
    title = "By Stress Level",
    x = "Stress Level",
    y = "Predicted Probability",
    color = "Disorder",
    fill = "Disorder"
  ) +
  theme_minimal(base_size = 14)

# 2. Predicted probabilities by Physical Activity Level
pred_probs_activity <- ggeffect(multi_logit, terms = "physicalActivityLevel")

p2 <- ggplot(pred_probs_activity, aes(x = x, y = predicted, color = group, fill = group)) +
  geom_line(size = 1.2) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.15, color = NA) +
  labs(
    title = "By Physical Activity Level",
    x = "Physical Activity Level",
    y = "Predicted Probability",
    color = "Disorder",
    fill = "Disorder"
  ) +
  theme_minimal(base_size = 14)

# Combine the plots side by side
combined_plot <- p1 + p2 +
  plot_annotation(title = "Predicted Probability of Sleep Disorder Types")

# Print the combined plot
print(combined_plot)
```


## 8: Modeling Sleep Quality with Ordered Logistic Regression
```{r}
library(scales)
library(broom)

# Model of the Ordered Logistic Regresion
sleepQualityOrderLogitRegression <- polr(qualityOfSleep ~ age + bmiCategory + dailySteps + occupation + gender, 
                                              data=sleepOLR, Hess=TRUE)
summary(sleepQualityOrderLogitRegression)

# Model Visualization
## Tidy the model
sleepQualityOLR_tidy <- broom::tidy(sleepQualityOrderLogitRegression)
sleepQualityOLR_tidy <- sleepQualityOLR_tidy[!grepl("\\|", sleepQualityOLR_tidy$term), ]

## Plot
ggplot(sleepQualityOLR_tidy, aes(x = estimate, y = reorder(term, estimate))) +
  geom_point() +
  geom_errorbarh(aes(xmin = estimate - std.error, xmax = estimate + std.error), height = 0.2) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Ordered Logistic Regression Coefficients",
       x = "Coefficient Estimate",
       y = "Predictor") +
  theme_minimal()
```


The ordered logistic regression model evaluates the relationship between age, BMI category, daily steps, occupation, and gender with sleep quality. Age shows a strong positive association with better sleep quality (estimate = 4.47, t = 11.96), suggesting that older individuals are significantly more likely to report higher sleep quality levels. In contrast, higher daily steps are modestly associated with lower sleep quality (estimate = –0.35, t = –2.12), though the effect size is small. BMI categories show strong negative impacts: being overweight (estimate = –7.03, t = –9.55) or obese (estimate = –5.43, t = –5.35) is significantly associated with worse sleep quality compared to the baseline BMI group, while normal weight shows no significant difference (t = –0.98). Occupation also plays a role; engineers (estimate = –1.95, t = –2.24), nurses (estimate = –2.02, t = –2.88), and salespeople (estimate = –3.19, t = –2.91) are notably more likely to report poorer sleep compared to the reference occupation. Gender, specifically being male, shows a positive but statistically insignificant effect on sleep quality (estimate = 1.05, t = 1.41). Overall, age and BMI are the strongest predictors in this model, with daily steps and certain occupations also contributing meaningfully to variations in reported sleep quality.


## 9: Advanced Topics
```{r}
# PCA
pca_input <- sleepPCA %>% select(-sleepDisorder)

pca_result <- prcomp(pca_input, center = TRUE, scale. = TRUE)

summary(pca_result)

# Clustering
# Calculate % variance explained for PC1 and PC2
pc_var <- summary(pca_result)$importance[2, 1:2]
pc1_var <- round(pc_var[1] * 100, 1)
pc2_var <- round(pc_var[2] * 100, 1)

pca_scores <- as.data.frame(pca_result$x[, 1:2]) # First two principal components

print("Estimating PCA Categories")
pca_result$rotation[, 1:2]

kmeans_result <- kmeans(pca_scores, centers = 2, nstart = 25)

# Add cluster labels to PCA scores
pca_scores$cluster <- as.factor(kmeans_result$cluster)

# Make sure labels are numeric
true_labels <- as.numeric(sleepPCA$sleepDisorder)
predicted_clusters <- as.numeric(kmeans_result$cluster)

# ARI: Adjusted Rand Index
ari_score <- adjustedRandIndex(true_labels, predicted_clusters)

# AMI: Adjusted Mutual Information
ami_score <- AMI(true_labels, predicted_clusters)

# Print results
print(paste("Adjusted Rand Index (ARI):", round(ari_score, 3)))
print(paste("Adjusted Mutual Information (AMI):", round(ami_score, 3)))
```

## 10: Advanced Topic Visualization
```{r}
factoextra::fviz_eig(pca_result)

# Plot WSS (Elbow method) to see ideal K for k-means clustering
fviz_nbclust(pca_scores[, 1:2], kmeans, method = "wss") +
  labs(title = "Elbow Method for Optimal k (PCA-Reduced Features)")

# Improved PCA Cluster Plot
ggplot(pca_scores, aes(x = PC1, y = PC2, color = cluster)) +
  geom_point(size = 2.5, alpha = 0.8) +
  labs(
    title = "K-Means Clustering of Sleep Patterns",
    x = paste0("Principal Component 1 (", pc1_var, "% variance)"),
    y = paste0("Principal Component 2 (", pc2_var, "% variance)"),
    color = "Cluster"
  ) +
  theme_minimal(base_size = 14) +
  scale_color_brewer(palette = "Set2")
```

## 11: Conclusion

Linear and Logistic Regression Conclusion:

Clustering and PCA conclusion:

The PCA results demonstrated that roughly two components can explain 94% of the variation in the data (between the three variables physicalActivityLevel, stressLevel, sleepDuration). The result of the PCA was plotted using the Scree plot where we see dimension 3 is significantly lower than dimensions 1 and 2. Using the rotation matrix of the pca results, we see that PC1 is likely quality of sleep, since it is most explained by sleep duration, and PC2 is likely physical activity level, since it is most explained by physicalActivityLevel. Since the Adjusted Rand Index and Adjusted Mutual Information are relatively low, the clusters derived from behavioral data do not strongly align with diagnosed sleep disorders.